Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-29 MS_thesis.tex
Rephrase after proofreading

%The abstract goes here. IISER guidelines say it should be <= 250 words.
\thesisabstract{
Population biology is built on a strong mathematical foundation developed during the Modern Synthesis. Historically, much of this foundation has been built using models of infinite populations, ignoring the effects of demographic stochasticity. Finite population models in population and quantitative genetics usually assume a fixed population size and are of limited applicability in a world where population sizes  routinely fluctuate. In this thesis, I analytically describe evolution in finite, fluctuating populations from biological first principles. Starting from a density-dependent `birth-death process’ describing an arbitrary closed population of individuals with discrete traits, I derive stochastic differential equations (SDEs) for how trait frequencies change over time. These SDEs reveal a new directional evolutionary force, `noise-induced selection’, that is particular to finite, fluctuating populations and has been largely overlooked in standard formulations of evolution. Noise-induced selection can cause deviations from neutrality and can reverse the direction of evolution predicted by infinite-population frameworks. My SDEs also recover well-known results such as the replicator-mutator equation and the Price equation in the infinite population limit and can help explain and organize several previous studies under a single theoretical framework. Finally, I extend the birth-death formalism to one-dimensional quantitative traits through a `stochastic field theory’ that yields frameworks such as Kimura's continuum-of-alleles and Lande's gradient dynamics in the infinite population limit. My work thus generalizes the formal structures of population biology to finite, fluctuating populations and predicts a new evolutionary force unique to such populations. I also discuss the implications for evolutionary agent-based models and empirical studies.
}

@@

The support and the resources provided by the PARAM Brahma Facility under the National
Supercomputing Mission, Government of India at the Indian Institute of Science Education
and Research (IISER), Pune are gratefully acknowledged; However, it is worth explicitly noting that while I did briefly run some scripts on the PARAM-Brahma facility that were related to my thesis project, none of the scripts or simulations run on the PARAM-Brahma supercomputer ultimately contributed in any way, shape, or form to any material presented in this thesis --- The main text is completely analytical, and Appendix \ref{App_examples} only uses very simple simulations that were run on my personal laptop; The particular simulations I ran on PARAM also did not help me gain any insight or understanding towards any of the phenomena and/or models studied in this thesis.
}

>>>>

%The abstract goes here. IISER guidelines say it should be <= 250 words.
\thesisabstract{
Population biology is built on a strong mathematical foundation developed during the Modern Synthesis. Historically, much of this foundation has been built using models of infinite populations, ignoring the effects of demographic stochasticity. Finite population models in population and quantitative genetics usually assume a fixed population size and are of limited applicability in a world where population sizes  routinely fluctuate. In this thesis, I analytically describe evolution in finite, fluctuating populations from biological first principles. Starting from a density-dependent `birth-death process’ describing an arbitrary closed population of individuals with discrete traits, I derive stochastic differential equations (SDEs) for how trait frequencies change over time. These SDEs reveal a new directional evolutionary force, `noise-induced selection’, that is particular to finite, fluctuating populations and has been largely overlooked in standard formulations of evolution. Noise-induced selection can cause deviations from neutrality and may reverse the direction of evolution predicted by infinite-population frameworks. My SDEs also recover well-known results such as the replicator-mutator equation and the Price equation in the infinite population limit, and help explain and organize several previous studies under a single theoretical framework. Finally, I extend the birth-death formalism to one-dimensional quantitative traits through a `stochastic field theory’ that yields frameworks such as Kimura's continuum-of-alleles and Lande's gradient dynamics in the infinite population limit. My work thus generalizes the formal structures of population biology to finite, fluctuating populations and predicts a new evolutionary force unique to such populations. I also discuss the implications for designing evolutionary agent-based models and empirical studies.
}

@@

The support and the resources provided by the PARAM Brahma Facility under the National
Supercomputing Mission, Government of India at the Indian Institute of Science Education
and Research (IISER), Pune are gratefully acknowledged; However, it is worth explicitly noting that while I did briefly run some scripts on the PARAM-Brahma facility that were related to my thesis project, none of the scripts or simulations run on the PARAM-Brahma supercomputer ultimately contributed to this thesis --- The main text is completely analytical, and Appendix \ref{App_examples} only uses very simple simulations that were run on my personal laptop.
}
